The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-08-22 paper.qmd
read over + edited background

From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming. In this overflow of information, tools that can effectively summarize information down into simple and clear ideas become more valuable. Information visualisations remain one of the most powerful tools for fast and reliable science communication.

Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set. Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] highlight this power in visualisation. Additionally, visualisations allow for efficient and memorable communication. Even something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].

Uncertainty visualisation is a relatively new field in research. Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992]. These early experiments typically involved showing participants a distribution, such as those depicted in @fig-ibrekk, and asking the viewers to extract a probability or average. Despite the new terminology visualisation of uncertainty has been present since the earliest times. For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution. Today, there is an abundance of publications on the topic which makes it is timely to construct a review of the field. In fact, there have already been several reviews published.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ibrekk
#| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments. This early experiment is a good example of many of the issues that are still common in uncertainty visualisation today. For example, the '95% confidence interval' is more accurately a '95% prediction interval'. Additionally graphics that depict different mathematical objects that are also different on their visual components are compared because of a percieved relation to uncertainty. So visualisations of the mean, PDF and CDF, are all discussed as though they all contain the relevant statistical information. The axis also have different scales, and visualisation methods that are now unpopular for displaying proportions, such as a pie chart, are used."
#| fig-subcap: 
#|   - "Picture 1"
#|   - "Picture 2"
#|   - "Picture 3"
#|   - "Picture 4"
#|   - "Picture 5"
#|   - "Picture 6"
#|   - "Picture 7"
#|   - "Picture 8"
#|   - "Picture 9"
#| layout-ncol: 3

@@

Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather comment on the *difficulties* faced when trying to summarize the field. @Kinkeldey2014 found most experiments on the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals. @Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic. @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances. @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of a collection of papers. @uncertchap2022 summarized several cognitive effects that have repeatedly arised in uncertainty literature, however these effects were each discussed in isolation as a list of considerations an author might make. While these reviews are thorough in scope, none discuss how the existing literature contribute to the commonly state goal of uncertainty visualisation, scientific transparency. The problem faced by the literature is easily summarized with a famous quote by Henri Poincaré.

@@

That is to say, despite the wealth of reviews, the field of uncertainty visualisation remains a heap of stones. This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem, and hopefully laying the foundations on which we can build a house.

This review is broken into several parts that each reflect a different approach to uncertainty visualisation. First we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all. Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem. Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss if the visualisations created by these transformations are still "uncertainty visualisations". Fourth, we will discuss methods that implicitly include uncertainty by depicting a sample or original data in place of an estimate. Finally, we discuss how uncertainty visualisations can be effectively evaluated. When discussing each of these methods, we consider the *purpose* of uncertainty visualisation and comment on how effective each visualisation is at fulfilling that purpose.

## Spatial examples
There is an endless number of uncertainty visualisations and it would be impossible to show all of them in this review. While the theoretical approach we outline is useful to all uncertainty visualisations, regardless of their application, we feel that showing the variations on a single visualisation will help isolate the ideas we are trying to convey. Due to the fields origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. We will use these maps to highlight the costs and benefits of each approach.

@fig-data shows the first six rows of our example data, along with the geographical boundaries of our data set. The temperature variable was generated using a quadratic spatial trend, that is $Temperature = 28-SLongitude^2 + SLatitude^2$ where the $SLongitide$ and $SLatitude$ are the scaled (to standard normal) longitude and latitude of the county's centroid. The high and low cases for the sampling variance are randomly sampled from a uniform distribution. The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard error cases are drawn from a a $U_{[1,2]}$ distribution. As we are dealing with an average, the sampling distribution would be normally distributed, so each county temperature estimate is assumed to come from a $N(Temp_i, SE_{case,i})$. This is the data we will be using in our spatial uncertainty examples for the rest of the paper.

@@

## What is uncertainty

@@

# Visualising uncertainty and signal in a new space
Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, and uncertainty visualisation authors already have.

>>>>

From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming. In the face of this overflow of information, tools that effectively summarize information down into simple and clear ideas become more valuable.

Among these summary tools, information visualisations are one of the most powerful. Visualisations allow us to **learn** what is important about our data set and datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] highlight this power. Additionally, visualisations allow for efficient and memorable communication. Even something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].

Uncertainty visualisation is a relatively new field in research. Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992]. @fig-ibrekk depicts an example of the uncertainty visualisations discussed in these early papers. Despite the new terminology visualisation of uncertainty has been present since the earliest times. For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution. Today, there is an abundance of publications on the topic which makes it is timely to construct a review of the field. In fact, there have already been several reviews published.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ibrekk
#| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments. Several visualisation methods that are now unpopular (such as the pie chart) are used throughout this paper."
#| fig-subcap: 
#|   - "Picture 1"
#|   - "Picture 2"
#|   - "Picture 3"
#|   - "Picture 4"
#|   - "Picture 5"
#|   - "Picture 6"
#|   - "Picture 7"
#|   - "Picture 8"
#|   - "Picture 9"
#| layout-ncol: 3

@@

Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather, they comment on the *difficulties* faced when trying to summarize the field. @Kinkeldey2014 found most experimental methods to be ad hoc, with no commonly agreed upon methodology, formalisations, or greater goal of describing general principals. @Hullman2016 noticed there is a serious noise issue in the data coming from uncertainty visualisation experiments. She commented on the prevalence of confounding variables that make it unclear as to what exactly caused a subjects poor performance on a particular questions. Mistakes due to misunderstood a visualisations, misinterpreted questions, and incorrectly applied heuristics are all combined into a single error value. @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances. @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of a collection of papers. @uncertchap2022 summarized several cognitive effects that repeatedly arise in uncertainty visualisation experiments, however these effects were each discussed in isolation as a list of considerations an author might make rather than an overarching theory of rules for effective uncertainty visualisation. While these reviews are thorough in scope, none discuss how the existing literature contribute to the commonly stated goal of uncertainty visualisation, scientific transparency. The problem faced by the literature is easily summarized with a famous quote by Henri Poincaré.

@@

That is to say, despite the wealth of reviews, the field of uncertainty visualisation remains a heap of stones. Lovely stones, but stones nonetheless. This review attempts to address this issue by offering a novel perspective on the uncertainty visualisation problem, and hopefully laying the foundations on which we can build a house.

This review is broken into several parts that each reflect a different approach to uncertainty visualisation. First we look at graphics that ignore uncertainty entirely and discuss why uncertainty should be included at all. Second, we look at methods that consider uncertainty to be just another variable and discuss the characteristics of uncertainty that make it a unique visualisation problem. Third, we look at methods that explicitly combine our estimate and its uncertainty and discuss the limitations of these approaches. Fourth, we will discuss methods that implicitly include uncertainty by depicting a sample in place of an estimate. Finally, we discuss how uncertainty visualisations can be effectively evaluated. When discussing each of these methods, we will repeatedly return to the *purpose* of uncertainty visualisation and the effectiveness of each approach in fulfilling tha purpose.

## Spatial example
It would be impossible to display every single uncertainty visualisation in this review, there are far too many so for the sake of clarity, we will focus on the changes made to a single plot. The choropleth map. Due to the fields origins and focus in geospatial information visualisation, there have been a large number of suggested variations on the choropleth map that allow authors to include uncertainty. While the theoretical approach we outline is useful to all uncertainty visualisations, regardless of their application, we feel that focusing on a single example will help isolate the ideas we are trying to convey. We will use these maps to show the effects we are talking about, and highlight the costs and benefits of each approach. These examples will need

@fig-data shows the first six rows and the geographical boundaries of our data set. The temperature variable was generated using a quadratic spatial trend, that is $Temperature_i = 28-Longitude_i^2 + Latitude_i^2$ where the $Longitide$ and $Latitude$ are the longitude and latitude of the county's centroid scaled to a standard normal distribution. Each county's variance is independently, randomly sampled from a uniform distribution. The low standard errors are drawn from a $U_{[0,1]}$ distribution, while the high standard errors are drawn from a a $U_{[1,2]}$ distribution. As we are dealing with an average, the sampling distribution would be approximately normal by the central limit theorem, so each county temperature estimate is assumed to come from a $N(Temp_i, SE_{case,i})$. This is the data we will be using in our spatial uncertainty examples for the rest of the paper.

@@

@@

# Combining uncertainty and signal in a transformed space
Instead of hoping that uncertainty might collapse signal values into a single dimension, we can do some of that work ourselves, and uncertainty visualisation authors already have.
